Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by Galaxy Cluster Abell 2744

### 2015-03-23 TABLE/magnifications.tex
incorporating most of Marusa's comments.   * including adding in a section on redshift error

\begin{deluxetable*}{lccp{1.5in}p{2.2in}}
\tablecolumns{5}
\tablecaption{Predicted magnifications for SN \tomas\ from lens models.\label{tab:PredictedMagnifications}}
\tablehead{ \colhead{Model} & \colhead{Best}\tablenotemark{a} & \colhead{Median\tablenotemark{b}} & \colhead{References} & \colhead{Description}}
\startdata
%Sharon         & 2.57$^{+0.19}_{-0.16}$ &   \citealt{Jullo:2007} \\[2mm]
Sharon(v2)   & 2.73 & 2.69$^{+0.14}_{-0.06}$ &   \citealt{Jullo:2007};\citealt{Johnson:2014} & LENSTOOL parametric, strong-lensing based model\\
CATS-SL      & 2.25 & 2.27$^{+0.05}_{-0.04}$ &   \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL hybrid parametric/free-form, strong lensing based model.\\
CATS-SL+WL   & \nodata & 2.62$^{+0.18}_{-0.18}$ & \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL hybrid parametric/free-form, strong+weak lensing based model.\\
Jauzac		 & \nodata & 3.37$^{+0.14}_{-0.15}$ &   \citealt{Jauzac:2014,Richard:2014} & Updated version of the CATS-SL model, adds 33 new multiply-imaged galaxies.\\
GLAFIC       & 2.32 & 2.28$^{+0.07}_{-0.11}$ &   \citealt{Oguri:2010,Ishigaki:2015} & Parametric strong-lensing model using the {\tt GLAFIC} code.\tablenotemark{d} \\
Zitrin-NFW   & 2.07 & 2.27$^{+0.23}_{-0.22}$ &   \citealt{Zitrin:2009a} &  Parametric strong-lensing model using PIEMD\tablenotemark{e} profiles for galaxies and NFW\tablenotemark{f} profiles for dark matter halos.\\
Zitrin-LTM   & 2.64 & 2.96$^{+0.77}_{-0.38}$ &   \citealt{Zitrin:2013a} & Parametric strong-lensing model, adopts the Light-Traces-Mass assumption for both the luminous and dark matter.\\
Bradac       & 3.15 & 2.45$^{+0.19}_{-0.16}$ &   \citealt{Bradac:2005,Bradac:2009} & Free-form, strong+weak-lensing based model\\
Williams     & 2.67 & 2.78$^{+2.68}_{-1.14}$ &   \citealt{Liesenborgs:2006,Mohammed:2014} & {\tt GRALE}\tablenotemark{f} : Free-form strong-lensing model using a genetic algorithm.  \\
Merten       & 2.31 & 2.22$^{+0.67}_{-0.19}$ &   \citealt{Merten:2009,Merten:2011} &  SaWLENS, free-form strong+weak lensing based model.\\
Lam			 & \nodata & 2.77$^{+0.36}_{-0.36}$ &   \citealt{Sendra:2014,Lam:2014} & {\tt WSLAP+}\tablenotemark{g} : Free-form strong+weak lensing model using a grid-based method, supplemented by deflections fixed to cluster member galaxies.
\enddata
\tablenotetext{a}{The magnification returned for the optimal version of each model, as independently defined by each lens modeling team.}  
\tablenotetext{b}{Median magnification from 100-600 Monte Carlo realizations of the model. Uncertainties enclose 68\%\ of the realized values.}  
\tablenotetext{c}{{\tt LENSTOOL} : \url{http://projects.lam.fr/repos/lenstool/wiki}}
\tablenotetext{e}{{\tt GLAFIC} : \url{http://www.slac.stanford.edu/~oguri/glafic/}}
\tablenotetext{e}{PIEMD: Pseudo Isothermal Elliptical Mass Distrubition}
\tablenotetext{f}{NFW : Navarro-Frenk-White mass density profile \citep{Navarro:1997}.}
\tablenotetext{g}{{\tt GRALE} : }
\tablenotetext{f}{{\tt WSLAP+} : Weak and Strong Lensing Analysis Package plus member galaxies}
\end{deluxetable*}

>>>>

\begin{deluxetable*}{lccp{1.5in}p{2.2in}}
\tablecolumns{5}
\tablecaption{Predicted magnifications for SN \tomas\ from lens models.\label{tab:PredictedMagnifications}}
\tablehead{ \colhead{Model} & \colhead{Best}\tablenotemark{a} & \colhead{Median\tablenotemark{b}} & \colhead{References} & \colhead{Description}}
\startdata
%Sharon         & 2.57$^{+0.19}_{-0.16}$ &   \citealt{Jullo:2007} \\[2mm]
Sharon(v2)   & 2.73 & 2.69$^{+0.14}_{-0.06}$ &   \citealt{Jullo:2007};\citealt{Johnson:2014} & LENSTOOL parametric, strong-lensing based model\\
CATS-SL      & 2.25 & 2.27$^{+0.05}_{-0.04}$ &   \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL hybrid parametric/free-form, strong lensing based model.\\
CATS-SL+WL   & \nodata & 2.62$^{+0.18}_{-0.18}$ & \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL hybrid parametric/free-form, strong+weak lensing based model.\\
Jauzac		 & \nodata & 3.37$^{+0.14}_{-0.15}$ &   \citealt{Jauzac:2014,Richard:2014} & Updated version of the CATS-SL model, adds 33 new multiply-imaged galaxies.\\
GLAFIC       & 2.32 & 2.28$^{+0.07}_{-0.11}$ &   \citealt{Oguri:2010,Ishigaki:2015} & Parametric strong-lensing model using the {\tt GLAFIC} code.\tablenotemark{d} \\
Zitrin-NFW   & 2.07 & 2.27$^{+0.23}_{-0.22}$ &   \citealt{Zitrin:2009a} &  Parametric strong-lensing model using PIEMD\tablenotemark{e} profiles for galaxies and NFW\tablenotemark{f} profiles for dark matter halos.\\
Zitrin-LTM   & 2.64 & 2.96$^{+0.77}_{-0.38}$ &   \citealt{Zitrin:2013a} & Parametric strong-lensing model, adopts the Light-Traces-Mass assumption for both the luminous and dark matter.\\
Bradac       & 3.15 & 2.45$^{+0.19}_{-0.16}$ &   \citealt{Bradac:2005,Bradac:2009} & {\tt SWUnited} : Free-form, strong+weak-lensing based model\\
Williams     & 2.67 & 2.78$^{+2.68}_{-1.14}$ &   \citealt{Liesenborgs:2006,Mohammed:2014} & {\tt GRALE}\tablenotemark{f} : Free-form strong-lensing model using a genetic algorithm.  \\
Merten       & 2.31 & 2.22$^{+0.67}_{-0.19}$ &   \citealt{Merten:2009,Merten:2011} &  SaWLENS, free-form strong+weak lensing based model.\\
Lam			 & \nodata & 2.77$^{+0.36}_{-0.36}$ &   \citealt{Sendra:2014,Lam:2014} & {\tt WSLAP+}\tablenotemark{g} : Free-form strong-lensing model using a grid-based method, supplemented by deflections fixed to cluster member galaxies.
\enddata
\tablenotetext{a}{The magnification returned for the optimal version of each model, as independently defined by each lens modeling team.}  
\tablenotetext{b}{Median magnification from 100-600 Monte Carlo realizations of the model. Uncertainties enclose 68\%\ of the realized values.}  
\tablenotetext{c}{{\tt LENSTOOL} : \url{http://projects.lam.fr/repos/lenstool/wiki}}
\tablenotetext{e}{{\tt GLAFIC} : \url{http://www.slac.stanford.edu/~oguri/glafic/}}
\tablenotetext{e}{PIEMD: Pseudo Isothermal Elliptical Mass Distrubition}
\tablenotetext{f}{NFW : Navarro-Frenk-White mass density profile \citep{Navarro:1997}.}
\tablenotetext{g}{{\tt GRALE} : }
\tablenotetext{f}{{\tt WSLAP+} : Weak and Strong Lensing Analysis Package plus member galaxies}
\end{deluxetable*}

### 2015-03-23 snTomas.tex
incorporating most of Marusa's comments.   * including adding in a section on redshift error

\begin{abstract}
SN \tomas\ is a Type Ia Supernova (\SNIa) discovered at $z=1.31$
behind the galaxy cluster Abell 2744 ($z=0.308$). This SN has a
projected separation from the cluster core of $\sim$40\arcsec, closer
than any other cluster-lensed \SNIa.  As such, it provides the first
opportunity to confront gravitational lens models of galaxy clusters
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  We derive a tightly constrained measure
of the peak apparent magnitude, corrected for light curve shape and
exinction.  In a cosmology-independent analysis, we find that \tomas\
is $0.75\pm0.10$ magnitudes brighter than unlensed \SNeIa\ at similar
redshift, implying a lensing magnification of $\mu_{\rm
obs}=2.00\pm0.19$.  We compare this direct measurement of the
magnification against predicted magnifications from 11 lens models,
representing a broad range of methodologies and incorporating a
variety of strong- and weak-lensing constraints.  The models are
collectively fairly accurate, with a naive unweighted mean of
$\mu_{\rm mod}=2.6$.  However, they are systematically biased to
higher values: none of the tested models predict a magnification equal
to or lower than the observed $\mu$, and some models are more than
$5\sigma$ above the measured magnification.  We evaluate possible
causes for this bias, and find that intercomparison of the lens models
rules out several simple possibilities.  A larger sample of
$\sim$10 \SNeIa\ behind any single cluster could provide a robust
evaluation of systematic biases in lens models.  Until such a sample
exists, it is advisable to increase the estimated uncertainty of any
predicted magnification by an amount comparable to the dispersion from
all available lens models.
\end{abstract}

@@

SN \tomas\ was discovered in Hubble Space Telescop (\HST) observations
with the Advanced Camera for Surveys (ACS) in the F606W and F814W
bands (V and i), collected on UT 2014 May 15 as part of the Hubble
Frontier Fields (HFF) survey (PI:J.Lotz,
HST-PID:13495).\footnote{\url{http://www.stsci.edu/hst/campaigns/frontier-fields}}
The HFF program is a 3-year director's discretionary initiative that
is collecting 140 orbits of HST imaging (roughly 340 ksec) on six
massive galaxy clusters, plus 6 accompanying ``parallel fields.''
Each field is observed in 3 optical bands (ACS F435W, F606W and F814W)
and 4 infrared (IR) bands (WFC3-IR F105W, F125W, F140W, and F160W),
although the optical and IR imaging campaigns are separated by $\sim$6
months. Abell 2744 was the first cluster observed, with IR imaging
spanning 2013 October-November, and optical imaging from 2014
May-July.  A composite image of the HFF data showing the SN is
presented in Figure~\ref{fig:DiscoveryImage}.  The SN detection was
made in difference images constructed using template imaging of Abell
2744 from HST+ACS observations taken in 2009 (PI:Dupke,
HST-PID:11689).

@@

The {\tt SNID} library can generate an acceptable \CCSN\ match only
when we remove all age and redshift constraints, In this case the best
non-Ia match is the Type Ic SN 1997ef, which delivers {\it rlap}$=6.8$
at $z=0.51$ and age=47.3 rest-frame days past peak.  This is not as
good a fit as the best Type Ia models, is at odds with the host galaxy
redshift prior, and is strongly disfavored by the shape and colors of
the SN light curve (see Section~\ref{sec:PhotometricClassification}).

@@

For Type Ia SNe, the SALT2 model has two additional parameters that
control the shape ($x_1$) and color ($c$) of the light curve.  We use
conservative priors here, defined to encompass a range of Type Ia SN
shapes and colors that is broader than typically allowed in
cosmological analyses \citep[see
e.g.][]{Kessler:2009b,Sullivan:2011,Rest:2014}. For $x_1$ the prior is
a bifurcated Gaussian distribution with mean $\bar{x}_1=0$, dispersion
$\sigma_{x_1}^+={0.9}$ and $\sigma_{x_1}^-={-1.5}$.  The prior for the
color parameter $c$ has $\bar{c}=0.0$, $\sigma_c^-=0.08$, and
$\sigma_c^+=0.54$.  The $c$ parameter in SALT2 combines intrinsic SN
color and extinction due to dust, so the large red tail of this
distribution allows for the possibility of several magnitudes
of dust extinction along the \tomas\ line of sight.

@@

This shift to higher magnifications is not, however, unique to the
SN \tomas\ position.  \citet{Jauzac:2014c} noted that their updated
mass model of Abell 2744 results in a systematic increase in the
magnification values across the cluster field. Examining a sample of
$\sim$30 multiple images, \citeauthor{Jauzac} found that the
magnifications increased by a factor of $\sim1.5$ relative to pre-HFF
models.  Although this single sight-line can only provide a very
limited test of the model, the $5.6\sigma$ discrepancy is at least a
strong suggestion that improving the number and quality of
strong-lensing constraints can still lead to magnification maps that
are susceptible to systematic biases.

@@

>>>>

\begin{abstract}
SN \tomas\ is a Type Ia Supernova (\SNIa) discovered at $z=1.31\pm0.01$
behind the galaxy cluster Abell 2744 ($z=0.308$). This SN has a
projected separation from the cluster core of $\sim$40\arcsec, closer
than any other cluster-lensed \SNIa.  As such, it provides the first
opportunity to confront gravitational lens models of galaxy clusters
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  We derive a tightly constrained measure
of the peak apparent magnitude, corrected for light curve shape and
exinction.  In a cosmology-independent analysis, we find that \tomas\
is $0.75\pm0.10$ magnitudes brighter than unlensed \SNeIa\ at similar
redshift, implying a lensing magnification of $\mu_{\rm
obs}=2.00\pm0.19$.  We compare this direct measurement of the
magnification against predicted magnifications from 11 lens models,
representing a broad range of methodologies and incorporating a
variety of strong- and weak-lensing constraints.  The models are
collectively fairly accurate, with a naive unweighted mean of
$\mu_{\rm mod}=2.6$.  However, they are systematically biased to
higher values: none of the tested models predict a magnification equal
to or lower than the observed $\mu$, and some models are more than
$5\sigma$ above the measured magnification.  We evaluate possible
causes for this bias, and find that intercomparison of the lens models
rules out several simple possibilities.  A larger sample of
$\sim$10 \SNeIa\ behind any single cluster could provide a robust
evaluation of systematic biases in lens models.  Until such a sample
exists, it is advisable to increase the estimated uncertainty of any
predicted magnification by an amount comparable to the dispersion from
all available lens models.
\end{abstract}

@@

SN \tomas\ was discovered in Hubble Space Telescop (\HST) observations
with the Advanced Camera for Surveys (ACS) in the F606W and F814W
bands (V and i), collected on UT 2014 May 15 as part of the Hubble
Frontier Fields (HFF) survey (PI:J.Lotz,
HST-PID:13495).\footnote{\url{http://www.stsci.edu/hst/campaigns/frontier-fields}}
The HFF program is a 3-year director's discretionary initiative that
is collecting 140 orbits of HST imaging (roughly 340 ksec) on six
massive galaxy clusters, plus 6 accompanying parallel fields.
Each field is observed in 3 optical bands (ACS F435W, F606W and F814W)
and 4 infrared (IR) bands (WFC3-IR F105W, F125W, F140W, and F160W),
although the optical and IR imaging campaigns are separated by $\sim$6
months. Abell 2744 was the first cluster observed, with IR imaging
spanning 2013 October-November, and optical imaging from 2014
May-July.  A composite image of the HFF data showing the SN is
presented in Figure~\ref{fig:DiscoveryImage}.  The SN detection was
made in difference images constructed using template imaging of Abell
2744 from HST+ACS observations taken in 2009 (PI:Dupke,
HST-PID:11689).

@@

The {\tt SNID} library can generate an acceptable \CCSN\ match only
when we remove all age and redshift constraints. In this case the best
non-Ia match is the Type Ic SN 1997ef, which delivers {\it rlap}$=6.8$
at $z=0.51$ and age=47.3 rest-frame days past peak.  This is not as
good a fit as the best Type Ia models, is at odds with the host galaxy
redshift prior, and is strongly disfavored by the shape and colors of
the SN light curve (see Section~\ref{sec:PhotometricClassification}).

@@

For Type Ia SNe, the SALT2 model has two additional parameters that
control the shape ($x_1$) and color ($c$) of the light curve.  We use
conservative priors here, defined to encompass a range of Type Ia SN
shapes and colors that is broader than typically allowed in
cosmological analyses \citep[see
e.g.,][]{Kessler:2009b,Sullivan:2011,Rest:2014}. For $x_1$ the prior is
a bifurcated Gaussian distribution with mean $\bar{x}_1=0$, dispersion
$\sigma_{x_1}^+={0.9}$ and $\sigma_{x_1}^-={-1.5}$.  The prior for the
color parameter $c$ has $\bar{c}=0.0$, $\sigma_c^-=0.08$, and
$\sigma_c^+=0.54$.  The $c$ parameter in SALT2 combines intrinsic SN
color and extinction due to dust, so the large red tail of this
distribution allows for the possibility of several magnitudes
of dust extinction along the \tomas\ line of sight.

@@

This shift to higher magnifications is not, however, unique to the
SN \tomas\ position.  \citet{Jauzac:2014c} noted that their updated
mass model of Abell 2744 results in a systematic increase in the
magnification values across the cluster field. Examining a sample of
$\sim$30 multiple images, \citeauthor{Jauzac:2014c} found that the
magnifications increased by a factor of $\sim1.5$ relative to pre-HFF
models.  Although this single sight-line can only provide a very
limited test of the model, the $5.6\sigma$ discrepancy is at least a
strong suggestion that improving the number and quality of
strong-lensing constraints can still lead to magnification maps that
are susceptible to systematic biases.

@@

\subsection{Redshift Error}
\label{sec:RedshiftError}

If the redshift of the SN derived in Section~\ref{sec:Spectroscopy}
were incorrect, then one would derive a different value for the
magnification, both from the SN measurement and the lens model
predictions.  Conceivably, this could resolve the tension between the
measurement and the models.  It is often the case in SN surveys that
redshifts are assigned based on a host galaxy association, typically
inferred from the projected separation between the SN and nearby
galaxies.  In this case the redshift evidence comes from the SN
itself, and we find a consistent redshift from both the SN spectrum
(Section~\ref{sec:SpecRedshift}) and the light curve
Section~\ref{sec:PhotometricClassification}, which are both within the
$1\sigma$ range of the photometric redshift for the nearest detected
galaxy: $z=1.5\pm0.2$. This appears to be a solid and self-consistent
picture, so the evidence strongly disfavors any redshift that is
significantly different from $z=1.3$.

We have adopted the spectroscopic redshift of $z=1.31\pm0.01$ for the
magnification comparison.  Spectroscopic redshifts are generally more
precise and accurate than those derived from photometric
observations \citep{Rodney:2010b,Kessler:2010,Sako:2011}.  However, in
this case the spectrum is limited to the rest-frame near-UV
wavelengths, where the available SN spectral template libraries are
more limited than at optical wavelengths.  If we use the photometric
redshift of $z=1.35\pm0.02$ instead, then the measured magnification
is reduced. \footnote{By increasing the redshift, the SN is presumed
to be at a greater distance, so to first order one would be expect it
to appear slightly fainter (larger distance modulus). If the distance
modulus inferred from the light curve fit remained constant, this
would drive the inferred magnification to a higher value.  However,
the intrinsic peak brightness as derived from light curve fitting is
more strongly affected by the covariance between redshift, color, and
light curve width.  In this case, these effects drive the distance
modulus to a larger value at $z=1.35$ that is closer to the unlensed
population than we get at $z=1.31$.}  From the SALT2 fitter we derive
$\mu_{\rm SALT2}=1.80\pm0.30$ and from MLCS2k2 we get $\mu_{\rm
MLCS2k2}=1.83\pm0.17$.  The lens models are also shifted, and they
uniformly move in the opposite direction, to magnification values that
are {\it larger}, by $\sim1$\%.  Thus, shifting the SN to the
photometric redshift of $z=1.35$ only serves to (slightly) exacerbate
the tension between the observations and models.
